Are populations like a circuit? Comparing isolation by resistance to a new coalescent-based method.

### 2019-05-03 isolation_by_coalescence.tex
final submission

% Responses to reviews:
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses
\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
\input{review-response-commands}
\newcommand{\responsefile}{review-responses.tex}  % name of the review reponses file

@@

\linenumbers

@@

Although commute times are uniquely determined by movement rates, the reverse is not true --
we can use equations \eqref{eqn:gh} and \eqref{eqn:R_from_H} to see how to modify movement rates
without changing commute times.  \llabel{ll:dense1}
A small concrete example of this is given in Appendix \ref{ss:simple_example}.
More generally, commute times only depend on the symmetric part of $H$ --
given a skew-symmetric matrix $Z$ (so that $Z + Z^T = 0$),
any Markov chain that has hitting times given by $H_\epsilon = H + \epsilon Z$ for some number $\epsilon$
will have the same commute times.
The resulting hitting times may not be valid --
if $G_\epsilon$ is the matrix constructed by applying equation \eqref{eqn:G_from_H} to $H_\epsilon$,
% (assuming this is invertible),
then it is not guaranteed that the offdiagonal elements of $G_\epsilon$ are all nonnegative, as required.
% (It is always true that rows of $G_\epsilon$ sum to zero.)
However, since $G$ and $H$ are continuous functions of each other,
if all offdiagonal elements of $G$ are strictly positive,
then there exists a positive $\epsilon_0$ such that all $G_\epsilon$ for $\epsilon < \epsilon_0$
do define valid Markov chains, all with the same commute times.

% \paragraph{Coalescence times}
The \emph{coalescence times} are defined using \emph{two} copies of the same chain,
as the mean time until coalescence,
if the chains coalesce at rate $\gamma$ when they are in the same place.
Concretely, 
suppose that $X$ and $Y$ are independent Markov chains both moving with movement rates given by $G$,
that coalesce at rate $\gamma_x$ when $X$ and $Y$ are both at $x$.
Define $\tau$ to be this coalescence time,
so that 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t = Y_t = x\} = \epsilon \gamma_x + O(\epsilon^2)$,
and 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t \neq Y_t\} = O(\epsilon^2)$.
Then, the (mean) coalescence time is defined to be
$C_{xy} = \E[\tau \given X_0 = x, \, Y_0 = y]$.
An argument similar to the one behind equation \eqref{eqn:hitting_sum}
shows that the coalescence times satisfy the following system of equations:
\begin{align}
\text{\bf (coalescence time)} \qquad
    \sum_y \left(G_{xy} C_{yz} + C_{xy} G_{zy}\right)
    &=
    \begin{cases}
        -1                   \quad & \text{if} \; x \neq z, \\
        -1 + \gamma_z C_{zz} \quad & \text{if} \; x = z.
    \end{cases}
\end{align}
Similar recursions are common in the literature,
going back at least to \citet{hill1972effective} (see also \citet{whitlock1997effective}).
In matrix notation, this is
\begin{align} \label{eqn:C_matrix}
    G C + C G^T - \diag(\gamma) \circ C = -\bone \bone^T,
\end{align}
where $\diag(\gamma)$ is the matrix with the vector $\gamma$ on the diagonal and zeros elsewhere,
and $\circ$ is the componentwise product.
In practice, we solve this by working with the product Markov chain $(X_t, Y_t)$,
whose generator matrix is $G \otimes I + I \otimes G$, \revpoint{1}{10}
where $I$ is the identity matrix and $\otimes$ is the Kronecker product.

@@

\paragraph{Forwards-time simulations}
To produce realistic data, we implemented forwards-time simulations using SLiM v3.1 
\citep{haller2018forward,haller2018treesequence},
with individuals living across continuous, two-dimensional geography (sometimes with barriers),
from which we recorded genomic data.
The basic simulation, which we modified to produce several other situations,
is as follows.
Each simulation had around 6400 diploid, hermaphroditic individuals, 
living across two-dimensional geography.
Mates were selected from nearby individuals using a Gaussian kernel with standard deviation $\sigma_d$,
truncated at $3 \sigma_d$.
Simulations were done on either a square region with width equal to $80\sigma_d$,
or a $5:3$ rectangular region with the same area.
Offspring are dispersed away from the mother using the same kernel,
reflected from the boundaries of the habitable area.
To produce a realistic model and prevent excessive population clumpiness \citep{felsensten1975pain}, 
we model local competition instead of a globally constant population size.
To do this, 
we use the same truncated Gaussian kernel to define an ``interaction strength''
between each pair of nearby individuals, 
and compute the probability of survival of each individual to the next time step as
$\min(0.9, 2 / (1 +  5x/12 \pi))$,
where $x$ is the sum of the interaction strengths of that individual with all other individuals
within distance $3 \sigma_d$.
(The value $x$ is an estimate of local population density.)
This reduces fitness for individuals in denser areas,
and produces an equilibrium population density
of close to one individual per $\sigma_d^2$ units of area.
The probability of survival for individuals within $10\sigma_d$ of the edge of the habitat
was multiplied by a linear factor that scaled from 0 to 1 with distance from the edge.
Patchiness of the resulting simulations was similar to what is seen in 
Figure \ref{fig:5x3b_continuous}.

Each diploid individual has one pair of homologous chromosomes with $10^8$ loci each 
and a recombination rate of $10^{-8}$ per generation per locus.
Neutral mutations were added at a rate of $10^{-9}$ per meiosis.
To make these simulations computationally feasible,
the mutation rate was set to 0.0 during the forwards simulation,
while tree sequences were recorded \citep{haller2018treesequence},
and mutations were added after the fact with msprime,
which is equivalent to including mutations during the forwards simulation 
but resulted in much faster run times \citep{kelleher2018efficient}.

@@

Thanks to David Levin for useful suggestions regarding hitting time calculations,
to Paul Marjoram for useful comments,
to Brad Shaffer and Evan McCartney-Melstad for discussions about landscape modeling,
to John Novembre, Hussein Al-Asadi and Benjamin Peter for help with EEMS
and general input,
and to Quentin Cronk and Armando Moreno Geraldes for consultations about the \emph{Populus} results.
Thanks also go to five anonymous reviewers for numerous useful suggestions.
We would also like to gratefully acknowledge the immense positive impact that Brad McRae \citep[1966--2017;][]{lawler2018tribute}
has had on landscape genetics and conservation biology through the introduction of landscape resistance.
We hope this paper helps to carry forward this work.
Work on this project was supported by funding from
the Sloan Foundation and the NSF (under DBI-1262645) to PR.

>>>>

% Responses to reviews:
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses
% \reviewresponsestrue  % include them
\reviewresponsesfalse  % don't include them
\input{review-response-commands}
\newcommand{\responsefile}{review-responses.tex}  % name of the review reponses file

@@

\nolinenumbers

@@

Although commute times are uniquely determined by movement rates, the reverse is not true --
we can use equations \eqref{eqn:gh} and \eqref{eqn:R_from_H} to see how to modify movement rates
without changing commute times.  \llabel{ll:dense1}
A small concrete example of this is given in Appendix \ref{ss:simple_example},
and depicted in Supplemental figure \ref{fig:valid_range}.
More generally, commute times only depend on the symmetric part of $H$ --
given a skew-symmetric matrix $Z$ (so that $Z + Z^T = 0$),
any Markov chain that has hitting times given by $H_\epsilon = H + \epsilon Z$ for some number $\epsilon$
will have the same commute times.
The resulting hitting times may not be valid --
if $G_\epsilon$ is the matrix constructed by applying equation \eqref{eqn:G_from_H} to $H_\epsilon$,
% (assuming this is invertible),
then it is not guaranteed that the offdiagonal elements of $G_\epsilon$ are all nonnegative, as required.
% (It is always true that rows of $G_\epsilon$ sum to zero.)
However, since $G$ and $H$ are continuous functions of each other,
if all offdiagonal elements of $G$ are strictly positive,
then there exists a positive $\epsilon_0$ such that all $G_\epsilon$ for $\epsilon < \epsilon_0$
do define valid Markov chains, all with the same commute times.

% \paragraph{Coalescence times}
The \emph{coalescence times} are defined using \emph{two} copies of the same chain,
as the mean time until coalescence,
if the chains coalesce at rate $\gamma$ when they are in the same place.
Concretely, 
suppose that $X$ and $Y$ are independent Markov chains both moving with movement rates given by $G$,
that coalesce at rate $\gamma_x$ when $X$ and $Y$ are both at $x$.
Define $\tau$ to be this coalescence time,
so that 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t = Y_t = x\} = \epsilon \gamma_x + O(\epsilon^2)$,
and 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t \neq Y_t\} = O(\epsilon^2)$.
Then, the (mean) coalescence time is defined to be
$C_{xy} = \E[\tau \given X_0 = x, \, Y_0 = y]$.
An argument similar to the one behind equation \eqref{eqn:hitting_sum}
shows that the coalescence times satisfy the following system of equations:
\begin{align}
\text{\bf (coalescence time)} \qquad
    \sum_y \left(G_{xy} C_{yz} + C_{xy} G_{zy}\right)
    &=
    \begin{cases}
        -1                   \quad & \text{if} \; x \neq z, \\
        -1 + \gamma_z C_{zz} \quad & \text{if} \; x = z.
    \end{cases}
\end{align}
Similar recursions are common in the literature,
going back at least to \citet{hill1972effective} 
(see also \citet{tufto1998analysis} and \citet{whitlock1997effective}).
In matrix notation, this is
\begin{align} \label{eqn:C_matrix}
    G C + C G^T - \diag(\gamma) \circ C = -\bone \bone^T,
\end{align}
where $\diag(\gamma)$ is the matrix with the vector $\gamma$ on the diagonal and zeros elsewhere,
and $\circ$ is the componentwise product.
In practice, we solve this by working with the product Markov chain $(X_t, Y_t)$,
whose generator matrix is $G \otimes I + I \otimes G$, \revpoint{1}{10}
where $I$ is the identity matrix and $\otimes$ is the Kronecker product.

@@

\paragraph{Forwards-time simulations}
To produce realistic data, we implemented forwards-time simulations using SLiM v3.1 
\citep{haller2018forward,haller2019treesequence},
with individuals living across continuous, two-dimensional geography (sometimes with barriers),
from which we recorded genomic data.
The basic simulation, which we modified to produce several other situations,
is as follows.
Each simulation had around 6400 diploid, hermaphroditic individuals, 
living across two-dimensional geography.
Mates were selected from nearby individuals using a Gaussian kernel with standard deviation $\sigma_d$,
truncated at $3 \sigma_d$.
Simulations were done on either a square region with width equal to $80\sigma_d$,
or a $5:3$ rectangular region with the same area.
Offspring are dispersed away from the mother using the same kernel,
reflected from the boundaries of the habitable area.
To produce a realistic model and prevent excessive population clumpiness \citep{felsensten1975pain}, 
we model local competition instead of a globally constant population size.
To do this, 
we use the same truncated Gaussian kernel to define an ``interaction strength''
between each pair of nearby individuals, 
and compute the probability of survival of each individual to the next time step as
$\min(0.9, 2 / (1 +  5x/12 \pi))$,
where $x$ is the sum of the interaction strengths of that individual with all other individuals
within distance $3 \sigma_d$.
(The value $x$ is an estimate of local population density.)
This reduces fitness for individuals in denser areas,
and produces an equilibrium population density
of close to one individual per $\sigma_d^2$ units of area.
The probability of survival for individuals within $10\sigma_d$ of the edge of the habitat
was multiplied by a linear factor that scaled from 0 to 1 with distance from the edge.
Patchiness of the resulting simulations was similar to what is seen in 
Figure \ref{fig:5x3b_continuous}.

Each diploid individual has one pair of homologous chromosomes with $10^8$ loci each 
and a recombination rate of $10^{-8}$ per generation per locus.
Neutral mutations were added at a rate of $10^{-9}$ per meiosis.
To make these simulations computationally feasible,
the mutation rate was set to 0.0 during the forwards simulation,
while tree sequences were recorded \citep{haller2019treesequence},
and mutations were added after the fact with msprime,
which is equivalent to including mutations during the forwards simulation 
but resulted in much faster run times \citep{kelleher2018efficient}.

@@

Thanks to David Levin for useful suggestions regarding hitting time calculations,
to Paul Marjoram and Nick Barton for useful comments,
to Brad Shaffer and Evan McCartney-Melstad for discussions about landscape modeling,
to John Novembre, Hussein Al-Asadi and Benjamin Peter for help with EEMS
and general input,
and to Quentin Cronk and Armando Moreno Geraldes for consultations about the \emph{Populus} results.
Thanks also go to five anonymous reviewers for numerous useful suggestions.
We would also like to gratefully acknowledge the immense positive impact that Brad McRae \citep[1966--2017;][]{lawler2018tribute}
has had on landscape genetics and conservation biology through the introduction of landscape resistance.
We hope this paper helps to carry forward this work.
Work on this project was supported by funding from
the Sloan Foundation and the NSF (under DBI-1262645) to PR.
